---
format: html
---

{{< include _load_data.qmd >}}

```{r}
#| label: report-libraries

library(FactoMineR)
library(factoextra)

```

```{r}
#| label: pca
candidates <- function(df) {
  n <- names(df)
  i2 <- length(n) - if ("geo_shape" %in% n) 2 else 0
  df |> select(all_of(16:i2)) |> select(where(\(col) !all(col == 0)))
}

top_n <- function(df, n) {
  sums <- map_int(df, sum) |> sort(decreasing = TRUE)
  top <- names(sums[1:n]) |> discard(is.na)
  select(df, all_of(top))
}

pcomps <- elections |>
  mutate(
    pco = map(data, \(df) prcomp(candidates(df) |> top_n(6), center = TRUE, scale. = TRUE))
  )

pcomps |>
filter(type == "presidentielles", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(fviz_screeplot(pco) + labs(title = str_glue("Élections {type} {year}, tour {round}")))
})

pcomps |>
filter(type == "presidentielles", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(
    fviz_pca_biplot(pco, label = "var", col.ind = "#BBBBBB", col.var="#0257df") +
    coord_fixed() +
    labs(title = str_glue("Élections {type} {year}, tour {round}"))
  )
})

```

```{r}
#| label: test-tmp

#elections$data |>
#  map_dfr(\(x) select(x, nb_votant, nb_procu, scrutin)) |>
#  mutate(scrutin = if_else(scrutin == "Législative", "Législatives", scrutin)) |>
#  ggplot(aes(x = nb_votant, y = nb_procu, color = scrutin)) +
#  geom_point(alpha = .2, size = .5)
```

### The Evolution of the Parisian Presidential Landscape (2007-2017-2022)

#### 2007
* **PC1 (The Classic Left/Right Divide):** The horizontal axis perfectly and diametrically opposes *Nicolas Sarkozy* (on the right side of the graph) to *Ségolène Royal* and *Olivier Besancenot* (on the left side). This is the historical left vs. right fracture of the 5th Republic.
* **PC2 (The Center vs. The Extremes):** The vertical axis is equally fascinating. It clearly isolates *François Bayrou* (the centrist) against the extreme of the political spectrum at the time, *Jean-Marie Le Pen*.

#### 2017
* **The Left/Right Divide Remains:** The horizontal axis still captures the traditional opposition between the Right (*François Fillon*) and the Left (*Jean-Luc Mélenchon* / *Benoît Hamon*).
* **The Explosion of PC2:** The vertical axis becomes massive and highlights a brand new fracture. It directly opposes *Emmanuel Macron* (pointing straight down) to *Marine Le Pen* (pointing straight up). This visually translates the emergence of a new "Progressivism/Globalism" vs. "Nationalism/Populism" divide that disrupted the old system.

#### 2022
* **The Fusion of the Right:** Look at *Valérie Pécresse* and *Éric Zemmour*. Their arrows are almost perfectly superimposed. Mathematically, this means that in Paris, the traditional conservative Right and the Far-Right appealed to the exact same sociology and neighborhoods (likely the affluent Western districts).
* **The Left Bloc:** Sitting on the complete opposite side of the *Pécresse*/*Zemmour* cluster is *Jean-Luc Mélenchon*, dominating his own half of the graph alone (representing the working-class/diverse Eastern districts).
* **The Central/Ecologist Bloc:** *Emmanuel Macron* and *Yannick Jadot* both point downwards in a similar general direction. In Paris, the Macron vote and the Ecologist vote in the first round share very similar sociological roots (the urban, upper-middle class).


```{r}
#elections |> pluck("data", 5) -> tmp
#candidates(tmp) -> cand
#prcomp(cand, center = TRUE, scale = TRUE) -> pc
#
#autoplot(pc, loadings = TRUE, loadings.label = TRUE, x = 1, y = 2)
#autoplot(pc, loadings = TRUE, loadings.label = TRUE, x = 1, y = 3)
#autoplot(pc, loadings = TRUE, loadings.label = TRUE, x = 2, y = 3)
#
#autoplot(pc, loadings = TRUE, loadings.label = TRUE, loadings.color = "blue", x = 1, y = 3) + coord_fixed()

```

```{r}
pcomps |>
filter(type == "europeennes", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(fviz_screeplot(pco) + labs(title = str_glue("Élections {type} {year}")))
})

pcomps |>
filter(type == "europeennes", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(
    fviz_pca_biplot(pco, label = "var", col.ind = "#BBBBBB", col.var="#0257df") +
    coord_fixed() +
    labs(title = str_glue("Élections {type} {year}"))
  )
})
```

### European Elections in Paris (2014 vs 2024)

#### 2014
* **PC1 (The Left/Right Spectrum):** The primary horizontal axis (explaining 55.5% of the variance) cleanly separates the traditional political families. On the left side of the graph, we find the conservative Right (*Alain Lamassoure*). On the right side, a scattered Left bloc (*Patrick Le Hyaric*, *Pascal Durand*, *Pervenche Berès*).

* In 2014, Paris still voted according to a highly fragmented, traditional multi-party system. No single candidate heavily dominated or merged with another's sociological base.

#### 2024
* **The Left/Right Shift:** The graph's orientation has flipped (which is mathematically normal in PCA), but the divide is starker. The Left bloc (*Aubry*, *Toussaint*, *Glucksmann*) is on the left side, while the Right/Center bloc (*Bellamy*, *Hayer*, *Bardella*) is on the right side.
* **The "Glucksmann/Toussaint" Fusion:**  Look at *Raphaël Glucksmann* (Center-Left) and *Marie Toussaint* (Greens). Their arrows are practically glued together, pointing to the exact same sociological space (down-left). In Paris, the moderate left and the ecologist electorate became virtually indistinguishable in 2024. This is the most striking finding of this chart.
* **A double polarization: ** We can clearly see a socio-economic axis (*Aubry* vs. *Bellamy* and *Hayer*) crossing with a cultural axis (*Glucksmann* and *Toussaint* vs. *Bardella*)

