---
title: Elections in Paris under the prism of SVD
format: html
---

{{< include _load_data.qmd >}}

```{r}
#| label: report-libraries

library(FactoMineR)
library(factoextra)
library(patchwork)
library(gt)

```

# Our Dataset

For this project, we will analyse the electoral results in Paris since 2007 (the year the official open data archives begin). Since a grand total of 16 elections happened since 2007, we only selected a subset that we found interesting:


| Election type| Years                  |
|--------------|------------------------|
| Presidential | 2007, 2017, 2022       |
| European     | 2014, 2024             |
| Legislative  | 2007, 2017, 2022, 2024 |
| Municipal    | 2008, 2020             |

For each of these election, we have a table where each individual is a polling station. The first 15 columns are informations about the election and the station (its location, number of registered voter, number of cast ballots...). Following this, there is one column per candidate which indicate the number of votes they received in each polling station.

As an example, here is a sample of what we can find about the first round of the 2017 presidential election:

```{r}
#| label: sample

elections |>
  filter(type == "presidentielles", year == 2017, round == 1) |>
  pluck("data", 1) |>
  select(!c(geo_shape, geo_point_2d)) |>
  rename_with(\(name) name |> str_replace_all("_", " ") |> str_to_title()) |>
  sample_n(5) |>
  gt() |>
  tab_header(title = "2017 Presidential elections, 1st round")

```



# Using Principal Component Analysis

TODO

```{r}
#| label: pca

candidates <- function(df) {
  n <- names(df)
  i2 <- length(n) - if ("geo_shape" %in% n) 2 else 0
  df |> select(all_of(16:i2)) |> select(where(\(col) !all(col == 0)))
}

top_n <- function(df, n) {
  sums <- map_int(df, sum) |> sort(decreasing = TRUE)
  top <- names(sums[1:n]) |> discard(is.na)
  select(df, all_of(top))
}

pcomps <- elections |>
  mutate(
    pco = map(data, \(df) prcomp(candidates(df) |> top_n(6), center = TRUE, scale. = TRUE))
  )

```

## The Evolution of the Parisian Presidential Landscape (2007-2017-2022)

```{r}
#| label: pca-pres

graphs <- pcomps |>
  filter(type == "presidentielles", round == 1) |>
  mutate(
    title = str_glue("{str_to_title(str_sub(type, end = 4))}. {year}, tour {round}"),
    screeplot = map(pco, partial(fviz_screeplot, addlabels = TRUE, main = title)),
    biplot = map(pco, partial(fviz_pca_biplot, label = "var", col.ind = "#BBBBBB", col.var="#0032b0", repel = TRUE, title = title))
  )

```

Let's first have a look at the screeplots of our PCA.

```{r}
#| label: pca-pres-screeplot

wrap_plots(graphs$screeplot, axes = "collect")

```

In every PCA, the first two principal axes contain at least 75% of the total variance. The elbow of the plot is not as pronounced as we usually want in a PCA but we can still neglect the 25% of variance remaining and look at the biplots:

```{r}
#| label: pca-pres-biplot

walk(graphs$biplot, \(x) print(x + coord_fixed()))

```

#### 2007
* **PC1 (The Classic Left/Right Divide):** The horizontal axis perfectly and diametrically opposes *Nicolas Sarkozy* (on the right side of the graph) to *Ségolène Royal* and *Olivier Besancenot* (on the left side). This is the historical left vs. right fracture of the 5th Republic.
* **PC2 (The Center vs. The Extremes):** The vertical axis is equally fascinating. It clearly isolates *François Bayrou* (the centrist) against the extreme of the political spectrum at the time, *Jean-Marie Le Pen*.

#### 2017
* **The Left/Right Divide Remains:** The horizontal axis still captures the traditional opposition between the Right (*François Fillon*) and the Left (*Jean-Luc Mélenchon* / *Benoît Hamon*).
* **The Explosion of PC2:** The vertical axis becomes massive and highlights a brand new fracture. It directly opposes *Emmanuel Macron* (pointing straight down) to *Marine Le Pen* (pointing straight up). This visually translates the emergence of a new "Progressivism/Globalism" vs. "Nationalism/Populism" divide that disrupted the old system.

#### 2022
* **The Fusion of the Right:** Look at *Valérie Pécresse* and *Éric Zemmour*. Their arrows are almost perfectly superimposed. Mathematically, this means that in Paris, the traditional conservative Right and the Far-Right appealed to the exact same sociology and neighborhoods (likely the affluent Western districts).
* **The Left Bloc:** Sitting on the complete opposite side of the *Pécresse*/*Zemmour* cluster is *Jean-Luc Mélenchon*, dominating his own half of the graph alone (representing the working-class/diverse Eastern districts).
* **The Central/Ecologist Bloc:** *Emmanuel Macron* and *Yannick Jadot* both point downwards in a similar general direction. In Paris, the Macron vote and the Ecologist vote in the first round share very similar sociological roots (the urban, upper-middle class).

```{r}
pcomps |>
filter(type == "europeennes", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(fviz_screeplot(pco) + labs(title = str_glue("Élections {type} {year}")))
})

pcomps |>
filter(type == "europeennes", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(
    fviz_pca_biplot(pco, label = "var", col.ind = "#BBBBBB", col.var="#0257df") +
    coord_fixed() +
    labs(title = str_glue("Élections {type} {year}"))
  )
})
```

## European Elections in Paris (2014 vs 2024)

#### 2014
* **PC1 (The Left/Right Spectrum):** The primary horizontal axis (explaining 55.5% of the variance) cleanly separates the traditional political families. On the left side of the graph, we find the conservative Right (*Alain Lamassoure*). On the right side, a scattered Left bloc (*Patrick Le Hyaric*, *Pascal Durand*, *Pervenche Berès*).

* In 2014, Paris still voted according to a highly fragmented, traditional multi-party system. No single candidate heavily dominated or merged with another's sociological base.

#### 2024
* **The Left/Right Shift:** The graph's orientation has flipped (which is mathematically normal in PCA), but the divide is starker. The Left bloc (*Aubry*, *Toussaint*, *Glucksmann*) is on the left side, while the Right/Center bloc (*Bellamy*, *Hayer*, *Bardella*) is on the right side.
* **The "Glucksmann/Toussaint" Fusion:**  Look at *Raphaël Glucksmann* (Center-Left) and *Marie Toussaint* (Greens). Their arrows are practically glued together, pointing to the exact same sociological space (down-left). In Paris, the moderate left and the ecologist electorate became virtually indistinguishable in 2024. This is the most striking finding of this chart.
* **A double polarization: ** We can clearly see a socio-economic axis (*Aubry* vs. *Bellamy* and *Hayer*) crossing with a cultural axis (*Glucksmann* and *Toussaint* vs. *Bardella*)
