---
title: Elections in Paris under the prism of SVD
format: html
---

{{< include _load_data.qmd >}}

```{r}
#| label: report-libraries

library(FactoMineR)
library(factoextra)
library(patchwork)
library(gt)
library(sf)

```

# Our Dataset

This project analyzes electoral results in Paris since 2007, the first year covered by the official open data archives. Because sixteen elections have taken place over that period, we focus on a curated subset that we find the most interesting for our analysis. The following table summarizes the elections we retained:


| Election type | Years                  |
|---------------|------------------------|
| Presidential  | 2007, 2017, 2022       |
| European      | 2014, 2024             |
| Legislative   | 2007, 2017, 2022, 2024 |
| Municipal     | 2008, 2020             |

These elections span the whole 2007-2022 period, which makes it possible to observe how voting patterns evolve over time. We also include every election from 2017 onward because the assignment requires it for the linear regression used to explain the 2024 legislative election.

For each election, we work with a table in which each row corresponds to a polling station. The first 15 columns contain information about the election and the station itself, such as its location, the number of registered voters, and the number of cast ballots. The remaining columns correspond to the candidates and record the number of votes each one received at that polling station.

The following example shows a sample of the data for the first round of the 2017 presidential election:

```{r}
#| label: sample

format_names <- function(name) name |> str_replace_all("_", " ") |> str_to_title()

elections |>
  filter(type == "presidentielles", year == 2017, round == 1) |>
  pluck("data", 1) |>
  select(!c(geo_shape, geo_point_2d)) |>
  rename_with(format_names) |>
  sample_n(5) |>
  gt() |>
  tab_header(title = "2017 Presidential elections, 1st round")

```

At first glance, the most popular candidates among Parisians were Emmanuel Macron, Jean-Luc Mélenchon, and François Fillon.

# Using Principal Component Analysis

In this section, we will perform PCA on the columns containing the results of the candidates for multiple elections and showcase some characteristics of the Parisian political landscape.

Since elections tend to have quite a lot of candidates obtaining only few votes, we focused our analysis on the 6 best scoring candidates of each election. We also ignore second rounds because having only two variables is not a very informative setup for PCA.

```{r}
#| label: pca

candidates <- function(df) {
  n <- names(df)
  i2 <- length(n) - if ("geo_shape" %in% n) 2 else 0
  df |> select(all_of(16:i2)) |> select(where(\(col) !all(col == 0)))
}

top_n <- function(df, n) {
  sums <- map_int(df, sum) |> sort(decreasing = TRUE)
  top <- names(sums[1:n]) |> discard(is.na)
  select(df, all_of(top))
}

pcomps <- elections |>
  mutate(
    pco = map(data, \(df) prcomp(candidates(df) |> top_n(6), scale. = TRUE)),
    title = str_glue("{str_to_title(str_sub(type, end = 4))}. {year}, tour {round}")
  )

```

## The Evolution of the Presidential Landscape (2007-2017-2022)

```{r}
#| label: pca-pres

pcomps.pres <- filter(pcomps, type == "presidentielles", round == 1)

graphs <- pcomps.pres |>
  mutate(
    screeplot = map(pco, partial(fviz_screeplot, addlabels = TRUE, main = title)),
    biplot = map(pco, partial(fviz_pca_var, col.var="royalblue", repel = TRUE, title = title))
  )

```

Let us start by the most well known election: the presidential election. For these elections, the top performing candidates are as follow:

```{r}
#| label: top-pres

pcomps.pres |>
  mutate(cand = map(pco, \(x) names(x$center) |> format_names())) |>
  select(Year = year, Candidates = cand) |>
  gt()
```

We first inspect the scree plots to check whether a two-dimensional representation is appropriate.

```{r}
#| label: pca-pres-screeplot

wrap_plots(graphs$screeplot, axes = "collect")

```

In every PCA, the first two principal axes explain at least 75% of the total variance. The elbow is not as sharp as one would usually hope for, but the remaining variance is still limited enough for us to focus on the correlation circles:

```{r}
#| label: pca-pres-biplot
#| lightbox:
#|  group: pca-pres-biplot

walk(graphs$biplot, print)

```

#### 2007
* **The classic left-right divide:** The horizontal axis clearly opposes *Nicolas Sarkozy* on the right with *Ségolène Royal* and *Olivier Besancenot* on the left. This reproduces the traditional left-right divide of the Fifth Republic.
* **The center vs. the extremes:** The vertical axis is equally interesting. It sets *François Bayrou*, the centrist candidate, against *Jean-Marie Le Pen*, who sat at the far end of the political spectrum at the time.

#### 2017
* **The left-right divide remains:** The horizontal axis still captures the traditional opposition between the Right (*François Fillon*) and the Left (*Jean-Luc Mélenchon* and *Benoît Hamon*).
* **Growing alternatives:** The vertical axis becomes more informative and reveals a second division. It directly opposes *Emmanuel Macron* (pointing downward) and *Marine Le Pen* (pointing upward), which corresponds to the emergence of a new divide between progressivism/globalism and nationalism/populism.

#### 2022
* **A narrowing on the right:** *Valérie Pécresse* and *Éric Zemmour* are almost perfectly superimposed. In practical terms, this suggests that in Paris the traditional conservative right and the far right appealed to very similar social profiles and neighborhoods, likely in the more affluent western districts.
* **The left bloc:** Sitting On the opposite side of the *Pécresse*/*Zemmour* cluster, *Jean-Luc Mélenchon* stands largely alone and dominates his half of the graph. This is consistent with support coming from more working-class and diverse eastern districts.
* **Macron's shift to the right:** Compared with the previous election, *Emmanuel Macron* now leans toward right-wing candidates. This suggests that perceptions of his politics shifted after his first term, and that his position in 2022 appeared more right-leaning than it did in 2017.

Since most of the political dynamic appears to be captured by the first principal axis, we now map each polling station according to its position on that axis.

```{r}
#| label: choropleth

pcomps.pres |>
  filter(year != 2007) |>
  mutate(
    choropleth = map2(data, pco, function(dt, pc) {
      bind_cols(dt, pc$x) |>
      st_as_sf() |>
      ggplot(aes(fill = PC1)) +
      geom_sf() +
      scale_fill_gradient2(low = "blue", high = "red") +
      labs(title = title)
    })
  ) |>
  pull(choropleth) |>
  wrap_plots(ncol = 1)
```

As expected, the map reveals a clear west-right / east-left divide in Paris. The western districts, especially the 16th and 17th arrondissements, are associated with more right-leaning values of the first principal component, while the eastern districts, especially the 19th and 20th, are associated with more left-leaning values. The first axis therefore captures a large part of the territorial opposition between affluent western neighborhoods and more popular eastern ones.

## European Elections (2014 vs 2024)

```{r}
pcomps |>
filter(type == "europeennes", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(fviz_screeplot(pco) + labs(title = str_glue("Élections {type} {year}")))
})

pcomps |>
filter(type == "europeennes", round == 1) |>
pwalk(function(type, year, round, pco, ...) {
  print(
    fviz_pca_biplot(pco, label = "var", col.ind = "#BBBBBB", col.var="#0257df") +
    coord_fixed() +
    labs(title = str_glue("Élections {type} {year}"))
  )
})
```

#### 2014
* **PC1 (The Left/Right Spectrum):** The primary horizontal axis (explaining 55.5% of the variance) cleanly separates the traditional political families. On the left side of the graph, we find the conservative Right (*Alain Lamassoure*). On the right side, a scattered Left bloc (*Patrick Le Hyaric*, *Pascal Durand*, *Pervenche Berès*).

* In 2014, Paris still voted according to a highly fragmented, traditional multi-party system. No single candidate heavily dominated or merged with another's sociological base.

#### 2024
* **The Left/Right Shift:** The graph's orientation has flipped (which is mathematically normal in PCA), but the divide is starker. The Left bloc (*Aubry*, *Toussaint*, *Glucksmann*) is on the left side, while the Right/Center bloc (*Bellamy*, *Hayer*, *Bardella*) is on the right side.
* **The "Glucksmann/Toussaint" Fusion:**  Look at *Raphaël Glucksmann* (Center-Left) and *Marie Toussaint* (Greens). Their arrows are practically glued together, pointing to the exact same sociological space (down-left). In Paris, the moderate left and the ecologist electorate became virtually indistinguishable in 2024. This is the most striking finding of this chart.
* **A double polarization: ** We can clearly see a socio-economic axis (*Aubry* vs. *Bellamy* and *Hayer*) crossing with a cultural axis (*Glucksmann* and *Toussaint* vs. *Bardella*)

# Clustering

TODO

# Linear regression

TODO
